Modal Verification Patterns for Systems Software

### 2025-09-13 paper.tex
reviews

In addition, the modal definitions themselves may represent ownership of resources in a substructural setting~\cite{restall1993modalities,dovsen1992modal,kamide2002kripke}. 
This allows modalities to simultaneously represent ownership of certain resources used for interpreting the modality itself, as well as whatever resource ownership is implied by the modal argument.
\subsection{Decomposing a System into its Constituents \emph{Contingently}}
\label{sec:definitions}
\begin{table*}[t]\small
\centering
\caption{Modal Decomposition of Program-Logics.}
\label{table:decomposition}
 %$\{\Diamond (P) \}$ write  $\{ P\}_{n}$
%$\{\overset{ICut^{n}}{\hookrightarrow} (P) \}$ return $\{ P\}_{n} $
\begin{tabular}{@{}lcccp{4cm}@{}}
  \toprule
{ Modality}& { Context} & { Elements}  &  { Nominalization} & { Context Steps} \\ \midrule
 Post-Crash~\cite{tejthesis,perennialgit,tejperennial19} &  $\Diamond \; P $  &  $  \ell \mapsto_{n}^{\overline{\gamma}} v $ &  Strong &   Crash Recovery    \\ 
%GC-Modality &      & \multicolumn{2}{c}{\multirow{2}{*}{Because why not}} & 8                                   & & ~\cite{}\\
 NextGen~\cite{larsnextgen25} &   $\overset{t}{\hookrightarrow} \; P$  &  Own (t(a))  &  Strong &  Determined Based on the Model$^{*}$   \\
 StackRegion$^{*}$  ~\cite{larsnextgen25} &  $\overset{ICut^{n}}{\hookrightarrow} \; P$      &  $\fbox{n} \; \ell \mapsto v$                                 &  Strong &  Alloc and Return to/from stack\\
 Memory-Fence~\cite{fsl,fsl++,derekrustbelt20} &  $\triangle_{\pi}$ and $\triangledown_{\pi}$      &  $\ell \mapsto v$                                 &  Weak &  Fence Acquire and Release  \\
 Address Space~\cite{kuru2025modalabstractionsvirtualizingmemory} &  [r]P     &   $\ell \mapsto v$ &  Weak                         &  Address Space Switch  \\ 
 Ref-Count~\cite{amalreal2024}& @$_{\ell}$ P    &  $\ell_1 \mapsto v$ &  Weak  &  Allocating, Dropping and Sharing a Reference \\
  Actor~\cite{gordon2019modal}& @$_{t}$ P    &  $(t,x,y)$ &  Strong  &  Send Message \\\bottomrule
\end{tabular}

@@

Except for Post-Crash-Modality, one can think of the resource contexts in the first column in Table \ref{table:decomposition} as containers for the corresponding resource elements in the second column.
\paragraph{Stack Regions} When reasoning about stack frame contents, the resource element would be a stack-memory points-to assertion ( $\fbox{n} \; \ell \mapsto v$)
indicating that a certain offset into stack region $n$ holds value $v$. 
%which is an instance of ownership assertion ($\textsf{Own(t(a))}$) for which a well-behaved transformation ($\textsf{t}$), e.g. $\textsf{ICut}^{n}$ algebra quantifying over valid stack region ids. 
\paragraph{Virtual Memory} For virtual memory management, a \textit{virtual-points-to}, $\ell \mapsto v$, is an ownership assertion pairing a virtual address ($\ell$) with data ($v$) in an address space, which is a resource context for virtual address mappings, $\ell \mapsto v$, and uniquely identified with a root address. Logically, specifying an address space switch ~\cite{kuru2025modalabstractionsvirtualizingmemory} from the address space with the root address $r2$ to the one with the root address $r1$
\[
\{[r1] P \ast Q \} cr3 := r1 \{P \ast [r2]Q\}_{\textsf{cr3}}
\]
requires having the resource elements loaded into the memory 
\[P \triangleq \ell_1 \mapsto v ...\]
to be in the resource context of the address space modality with the root $r1$, and, once the new address space is loaded—i.e., cr3 is loaded with the other address space root address—the resource elements ($Q$) that were loaded in the previous view of the memory are now introduced to the resource context with the root $r2$.

@@

\paragraph{Reference Counting} When dealing with reference counting APIs, we may care to specify reachability of memory nodes ($\ell \mapsto v$) in a certain context defined by a shared root address. A shared memory address $\ell$ can be the root of the graph ($@_{\ell}\; ( \ell_1 \mapsto v_1 \;\ast\; ... \; \ast \; \ell_j \mapsto v_j )$) that can be a container of memory nodes ($\ell_1 \mapsto v$) that are reachable from the root $\ell$. 
\paragraph{Actors} When dealing with reference counting APIs, we may care to specify reachability of memory nodes ($\ell \mapsto v$) in a certain context defined by a shared root address. A shared memory address $\ell$ can be the root of the graph ($@_{\ell}\; ( \ell_1 \mapsto v_1 \;\ast\; ... \; \ast \; \ell_j \mapsto v_j )$) that can be a container of memory nodes ($\ell_1 \mapsto v$) that are reachable from the root $\ell$. 
\paragraph{Post-Crash}
%\Colin{Ismail, not sure how to describe this. We don't want to go into the weeds on the NextGen}
%that can be jumped from a shared memory referenced whose reference count is at least 1 are also example resource elements from memory management task at different abstraction levels. 
The resource element of Post-Crash Modality is not obvious in Table \ref{table:decomposition}, and needs a bit of explanation. Perennial, based on the Iris logic, has both disk-points-to assertions $d[\textsf{p}] \mapsto_{n} v$ (for a specific disk $d$) and in-memory points-to assertions $\ell \mapsto_{n}^{\overline{\gamma}}$. Perennial crash-recovery logic book-keeps resource names (can be thought of as logical variables) $\overline{\gamma}$ to identify 
which assertions (resource elements) remain valid after a crash --- these assertions
are only usable while the names in $\overline{\gamma}$ are valid, and a crash resets
them, discarding assumptions about volatile state. A subtlety of the notion of a resource context is that, unlike the earlier examples, the context does not need to be a literal data structure. It can instead be (various forms of) a set of executions, as in the Post-Crash and NextGen modalities.
The Post-Crash modality $\diamond$ expresses that the assertion $P$ will be true after a crash discards
all unstable storage (i.e., RAM).
This was the inspiration for the NextGen modality, which is in fact a framework for defining ``after-$t$'' modalities, where $t$ is an transformation of the global state.\footnote{The transformations are subject to some
technical constraints that are unrelated to our point here.}
%whose capability is considered to be relevant to be kept in the crash invariant $\fbox{C}_n$, and to be used when a crash occurs.\Colin{Ismail, not sure about last line above. Can you better explain gamma?}
%regions of state (including auxiliary/ghost state) which persist after a crash (anything not identified

>>>>

In addition, the modal definitions themselves may represent ownership of resources in a substructural setting~\cite{restall1993modalities,dovsen1992modal,kamide2002kripke}. 
This allows modalities to simultaneously represent ownership of certain resources used for interpreting the modality itself, as well as whatever resource ownership is implied by the modal argument.
\subsection{Decomposing a System into its Constituents \emph{Contingently}}
\label{sec:definitions}
\begin{table*}[t]\small
\centering
\caption{Modal Decomposition of Program-Logics.}
\label{table:decomposition}
 %$\{\Diamond (P) \}$ write  $\{ P\}_{n}$
%$\{\overset{ICut^{n}}{\hookrightarrow} (P) \}$ return $\{ P\}_{n} $
\begin{tabular}{@{}lcccp{4cm}@{}}
  \toprule
{ Modality}& { Context} & { Elements}  &  { Nominalization} & { Context Steps} \\ \midrule
 Post-Crash~\cite{tejthesis,perennialgit,tejperennial19} &  $\Diamond \; P $  &  $  \ell \mapsto_{n}^{\overline{\gamma}} v $ &  Strong &   Crash Recovery    \\ 
%GC-Modality &      & \multicolumn{2}{c}{\multirow{2}{*}{Because why not}} & 8                                   & & ~\cite{}\\
 NextGen~\cite{larsnextgen25} &   $\overset{t}{\hookrightarrow} \; P$  &  Own (t(a))  &  Strong &  Determined Based on the Model$^{*}$   \\
 StackRegion$^{*}$  ~\cite{larsnextgen25} &  $\overset{ICut^{n}}{\hookrightarrow} \; P$      &  $\fbox{n} \; \ell \mapsto v$                                 &  Strong &  Alloc and Return to/from stack\\
 Actor~\cite{gordon2019modal}& @$_{t}$ P    &  $(t,x,y)$ &  Strong  &  Send Message \\
 Memory-Fence~\cite{fsl,fsl++,derekrustbelt20} &  $\triangle_{\pi}$ and $\triangledown_{\pi}$      &  $\ell \mapsto v$                                 &  Weak &  Fence Acquire and Release  \\
 Address Space~\cite{kuru2025modalabstractionsvirtualizingmemory} &  [r]P     &   $\ell \mapsto v$ &  Weak                         &  Address Space Switch  \\ 
 Ref-Count~\cite{amalreal2024}& @$_{\ell}$ P    &  $\ell_1 \mapsto v$ &  Weak  &  Allocating, Dropping and Sharing a Reference \\ \bottomrule
\end{tabular}

@@

Except for Post-Crash-Modality, one can think of the resource contexts in the first column in Table \ref{table:decomposition} as containers for the corresponding resource elements in the second column.
\paragraph{Stack Regions} When reasoning about stack frame contents, the resource element would be a stack-memory points-to assertion ( $\fbox{n} \; \ell \mapsto v$)
indicating that a certain offset into stack region $n$ holds value $v$. 
%which is an instance of ownership assertion ($\textsf{Own(t(a))}$) for which a well-behaved transformation ($\textsf{t}$), e.g. $\textsf{ICut}^{n}$ algebra quantifying over valid stack region ids. 
\paragraph{Virtual Memory} For virtual memory management, a \textit{virtual-points-to}, $\ell \mapsto v$, is an ownership assertion pairing a virtual address ($\ell$) with data ($v$) in an address space, which is a resource context for virtual address mappings, $\ell \mapsto v$, and uniquely identified with a root address. Logically, specifying an address space switch ~\cite{kuru2025modalabstractionsvirtualizingmemory} from the address space with the root address $r2$ to the one with the root address $r1$
\[
\{[r1] P \ast Q \} cr3 := r1 \{P \ast [r2]Q\}_{\textsf{cr3}}
\]
requires having the resource elements loaded into the memory 
\[P \triangleq \ell_1 \mapsto v ...\]
to be in the resource context of the address space modality with the root $r1$, and, once the new address space is loaded—i.e., cr3 is loaded with the other address space root address—the resource elements ($Q$) that were loaded in the previous view of the memory are now introduced to the resource context with the root $r2$.

@@

\paragraph{Reference Counting} When dealing with reference counting APIs, we may care to specify reachability of memory nodes ($\ell \mapsto v$) in a certain context defined by a shared root address. A shared memory address $\ell$ can be the root of the graph ($@_{\ell}\; ( \ell_1 \mapsto v_1 \;\ast\; ... \; \ast \; \ell_j \mapsto v_j )$) that can be a container of memory nodes ($\ell_1 \mapsto v$) that are reachable from the root $\ell$. 
\paragraph{Actors} When specifying the consistency of an actor in the current view,  we need to refer to other actors' state: $@_{t}(P)$ asserts that $P$ is true in the actor reference by a token identifying the message class. To be able to validate that the knowledge shared via immutable messages (resource elements), $(t,x,y)$, are preserved, the logic has to acknowledge the message class, $t$ both when referring to the messages and when referring to the other actors' state.   
\paragraph{Post-Crash}
%\Colin{Ismail, not sure how to describe this. We don't want to go into the weeds on the NextGen}
%that can be jumped from a shared memory referenced whose reference count is at least 1 are also example resource elements from memory management task at different abstraction levels. 
The resource element of Post-Crash Modality is not obvious in Table \ref{table:decomposition}, and needs a bit of explanation. Perennial, based on the Iris logic, has both disk-points-to assertions $d[\textsf{p}] \mapsto_{n} v$ (for a specific disk $d$) and in-memory points-to assertions $\ell \mapsto_{n}^{\overline{\gamma}}$. Perennial crash-recovery logic book-keeps resource names (can be thought of as logical variables) $\overline{\gamma}$ to identify 
which assertions (resource elements) remain valid after a crash --- these assertions
are only usable while the names in $\overline{\gamma}$ are valid, and a crash resets
them, discarding assumptions about volatile state. A subtlety of the notion of a resource context is that, unlike the earlier examples, the context does not need to be a literal data structure. It can instead be (various forms of) a set of executions, as in the Post-Crash and NextGen modalities.
The Post-Crash modality $\diamond$ expresses that the assertion $P$ will be true after a crash discards
all unstable storage (i.e., RAM).
This was the inspiration for the NextGen modality, which is in fact a framework for defining ``after-$t$'' modalities, where $t$ is an transformation of the global state.\footnote{The transformations are subject to some
technical constraints that are unrelated to our point here.}
%whose capability is considered to be relevant to be kept in the crash invariant $\fbox{C}_n$, and to be used when a crash occurs.\Colin{Ismail, not sure about last line above. Can you better explain gamma?}
%regions of state (including auxiliary/ghost state) which persist after a crash (anything not identified
